# Load Data {#load}

A forest inventory is a fundamental and comprehensive technical tool designed to collect as much information as possible about the status, composition, health, and productive potential of forest ecosystems.1 Unlike other assessment methods that may rely on remote sensing or large-scale modeling, field inventory is distinguished by its focus on direct data collection and in-situ measurements.2 This process not only generates detailed knowledge for sustainable management, but when carried out continuously in ten-year cycles, as in the Spanish National Forest Inventory, it allows the evolution of forest stands to be monitored over time.1 The methodology is based on sampling representative plots, which allows accurate estimates of key parameters to be obtained for vast geographical areas.2

We will use a data set published by Urquiza Muñoz et al,[@urquizamuñoz2021]. To start with, let's load and explore the data.

```{r load, message=FALSE, layout="l-body-outset"}
inventory <- read.csv("data/inventory_JDUM.csv")
```

In order to visualize the data in Rmarkdown, the following code will be employed:


```{r }
knitr::kable(head(inventory,10))
```

