# Chapter 1: Introduction to Forest Data in R {#Chapter 1: Introduction to Forest Data in R}

Welcome to the foundation of our forestry visualization journey. Before we can generate beautiful maps or 3D canopy models, we must first import and prepare our field data. R is uniquely powerful for this, as it allows us to document our data-cleaning steps transparently and reproducibly.

In this chapter, we will set up our R environment, import our raw forest inventory data, and perform basic cleaning to ensure it is ready for spatial and statistical analysis.

## 1.1 Setting up the Environment

To work effectively with forestry data, we need a specific set of tools. The tidyverse is a collection of R packages designed for data science (providing tools for reading, manipulating, and visualizing data), while sf (Simple Features) is the modern standard for handling spatial data in R.

Let's load our primary packages.

```{r}
#| label: setup
#| message: false
#| warning: false

# Install packages if you haven't already: install.packages(c("tidyverse", "sf", "knitr"))
library(tidyverse)
library(sf)
library(knitr) # Used for rendering neat tables in our book
```


## 1.2 Importing the Inventory Data

Forestry data collected in the field often comes in various formats. Our dataset is a CSV file, but like many datasets generated in South America or Europe, it uses a semicolon (;) to separate columns instead of a comma.

We must explicitly tell R how to read this specific format using the sep argument.

```{r}
#| label: load-data

# Import the raw data
raw_inventory <- read.csv("data/inventory.csv", 
                          sep = ";", 
                          dec = ".", 
                          stringsAsFactors = FALSE)

# Check the dimensions of our dataset (number of rows and columns)
dim(raw_inventory)
```
## 1.3 Data Cleaning and Standardization

Raw field data rarely comes perfectly formatted for analysis. Column names might include spaces, special characters, or abbreviations that are hard to remember.

Let's clean our dataset by renaming the most critical columns to standard English forestry terms and selecting only the variables we need for our upcoming visualizations.

```{r}
#| label: clean-data

inventory_clean <- raw_inventory %>%
  # Rename columns for easier coding later
  rename(
    Plot_ID = SP,
    Tree_ID = Nro.Arb,
    Common_Name = NOMBRE.VULGAR,
    Scientific_Name = NOMBRE.CIENTIFICO,
    DBH_cm = DAP,
    Height_m =HT,
    Easting = ESTE,
    Northing = NORTE
  ) %>%
  # Keep only the columns relevant for our visual analysis
  select(Plot_ID, Tree_ID, Common_Name, Scientific_Name, DBH_cm, Height_m,Easting, Northing) %>%
  # Filter out any rows that might be missing critical spatial or size data
  filter(!is.na(DBH_cm), !is.na(Easting), !is.na(Northing))
```

## 1.4 Exploring the Cleaned Data

Before moving on to complex visualizations, it is good practice to visually inspect the first few rows of our cleaned dataset to ensure our import and manipulation steps worked correctly.

```{r}
#| label: preview-data

# Display the first 5 rows as a polished table
kable(head(inventory_clean, 5), 
      caption = "Table 1.1: Preview of the cleaned forest inventory dataset.")
```


