# Forest Structural attributes

“Forest structure” usually refers to the way in which the attributes of trees are distributed within a forest ecosystem. 
Structure is a fundamental notion referring to patterns and relationships within a more or less well-defined system. We recognise structural attributes of buildings, crystals and proteins. Computer scientists design data structures and mathematicians analyse algebraic structures. All these tend to be rigid and complete. There are also open and dynamic structures.

Important theoretical concepts relating to biological structures include self-organisation, structure/property relations and pattern recognition. Self-organisation involves competition and a variety of interactions between individual trees.

The structure of a forest is the result of both natural processes and human disturbance. Important natural processes are species-specific tree growth, mortality and recruitment and natural disturbances such as fire, wind or snow damage. In
addition, human disturbance in the form of clear fellings, plantings or selective tree removal has a major structuring effect. The condition of the majority of forest ecosystems today is the result of human use (Sanderson et al. 2002; Kareiva et al. 2007).

This section introduces methods that can be used for describing structure based on tree diameters and diameter/height biomass relations.

## Summary  of strucutral attributes

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```


```{r message=FALSE, warning=FALSE}
inventory <- read.csv("data/inventory_JDUM.csv",sep=";")

table1=inventory %>%
  group_by(site) %>%
  summarise(n_trees = n(),
            dbh_mean = round(mean(dbh), 2),
            dbh_sd=round(sd(dbh), 1),
            ht_mean = round(mean(ht), 2),
            ht_sd=round(sd(ht), 1))
```

::: { .rmdnote}
_**Explaining the code**_:

**inventory**= _our dataset._

**%>%** = _this simbols are called pipe, and it this used to link the dataset to process._

**group_by()**= _this is used to group our summary based into a variable. In our example, we are using "site" to group the output summary by sites._

**summarise**= _function to summaries._

**n_tree, dbh_mean, ht_mean**= _names for our summary output, you can use any other name._

**n()**= _function to gives the current group size. In this case is related to number of trees._

**mean()**= _function to calculate the average of a variable. e.g: mean(dbh), mean(ht)._

**sd()**= _function to calculate the standard deviation of a variable. e.g: sd(dbh), sd(ht)._


**round()**= _rounds the values in its first argument to the specified number of decimal places (default 0). In our example the decimal places is two._
:::

```{r}
knitr::kable(table1)
```



## Diameter
Diameter at Breast Height (DBH): The DBH is the measurement of a tree's trunk diameter, taken at a standardized height of 1.30 meters from the ground level.5 In the United States, this measurement is taken at 4.5 feet (approximately 1.37 meters), while in Japan, it is defined as 1.25 meters.6 The standardization of this height is critical because it is chosen to avoid the influence of the natural swelling that occurs at the base of the trunk, which ensures the comparability of measurements between different trees and over time.6 The DBH is the most important variable in dendrometry, serving as a direct input for calculating basal area and as a fundamental variable in equations for estimating volume and biomass.3



###  Diametric classes

The breast height diameter of a forest tree is easy to measure and a frequently used variable for growth modeling, economic decision-making and silvicultural planning. Frequency distributions of tree diameters measured at breast height (DBH) are often available, providing a useful basis for an initial analysis of forest structure. ^[http://link.springer.com/10.1007/978-94-007-2202-6]

A grouping of tree diameters (DBH or DRC) into classes of a specified range. For some diameter classes, the number referenced (e.g., 4”, 6”, 8”) is designated as the midpoint of an individual class range. For example, if 2-inch classes are specified (the range for an individual class) and even numbers are referenced, the 6-inch class would include trees 5.0- to 6.9-inches in diameter.^[https://www.fs.fed.us/rm/ogden/publications/reference/periodic_terminology_final.pdf]

Formula used: Sturges rule to find the number of classes is given by:  

$$K = 1 + 3.322 * log(N)$$

Where K is the number of classes and N is the total frequency.

Example using Sturges rule in r (step by step)
```{r}
# range
R <- max(inventory$dbh)-min(inventory$dbh)
R

# intervals
K <- 1 + 3.322 * log(nrow(inventory))
K

# amplitude
A <- R / K
A
```

Now apply this to classes to each tree

An alternative method for the categorization of our DBH data into diameter classes is as follows:

```{r}
inventory$DC=trunc(inventory$dbh/10)*10
```

::: { .rmdnote}
_**Explaining the code**_:

**inventory$DC**= _create a new column named CD (diamtric class) into inventory dataframe._

**trunc**= _Is a function accepts a single numeric argument, designated as x, and produces a numeric vector containing the integers formed by truncating the values in x._
:::

plotting our results:

```{r}
inventory%>%
  ggplot(aes(x=DC))+
  geom_bar()+
  scale_x_continuous(breaks=seq(10,150,10))
```


### Diametric ditribution ^[http://link.springer.com/10.1007/978-94-007-2202-6]

Knowing the diameter distribution of trees is important for managing forests and predicting future timber yields. This information helps forest managers make informed decisions about harvesting, thinning, and conservation. The Weibull distribution is a common and flexible statistical model used to describe this distribution.

**Importance of Knowing Tree Diameter Distribution**

The distribution of tree diameters in a forest provides crucial insights into its structure, health, and productivity.


::: { .rmdnote}
_**Yield Prediction**: By knowing the distribution, foresters can accurately estimate the volume and value of timber available for harvest._

_**Forest Health Assessment**: A healthy, well-managed forest often has a specific diameter distribution. For example, a "reverse J-shaped" distribution (where there are many small trees and few large ones) is characteristic of a healthy, regenerating forest._

_**Management Decisions**: The distribution helps guide decisions on silvicultural treatments. If a stand has too many small, crowded trees, thinning may be necessary to promote the growth of the remaining trees._

_**Biodiversity and Habitat**: A diverse range of tree diameters supports a wider variety of wildlife habitats and promotes ecological resilience._
::: 


Diameter distribution probability curves were calculated using a Weibull distribution (de Lima et al., 2015, Rinne, 2008, Aigbe and Omokhua, 2014, Klaus von Gadow, 2012) based on all 5120 trees measured and classified by severity. 


#### Unimodal diameter frecuency

$$F(x)=P(X\le x)=\int_{- \infty }^{X} D(X)dX=1-e^{-(\frac {x}{scale})^{shape}}$$

**Weibull Distribution (adapted from Gadow, 2012):**

Create a summary table:

```{r}
DC_dist=inventory%>%
  group_by(DC)%>%
  summarise(n_tree=n_distinct(no1))
```


```{r warning=FALSE}
DC_dist2=NULL;
for(i in 1:length(DC_dist$DC)){DC_dist2=c(DC_dist2,rep(DC_dist$DC[i],times=DC_dist$n_tree[i]))}

hist(DC_dist2,freq=FALSE,ylim=c(0,0.1))


wu=MASS::fitdistr(DC_dist2,"weibull")

lines(10:90,dweibull(10:90,wu$e[1],scale=wu$e[2]),col="red")

```



## Tree Height


### Diameter-heigth relation

One of the most important elements of forest structure is the relationship between tree diameters and heights. Information about size-class distributions of the trees within a forest stand is important for estimating product yields.

```{r}
p=inventory %>%
  ggplot(aes(x = dbh, y = ht)) +
  geom_point() +
  labs(x = "DBH (cm)", y = "Tree height")

p
```

The regression line and equation can added as flow:
```{r}
library(ggpubr)

p+
geom_smooth(method='lm')+ 
stat_regline_equation(label.y = 50,size=5)+
stat_cor(label.y = 45,size=5)
```


## Basal area

The expression for the basal area (G) of an individual tree is the equation for a circle (expressed in m2 tree). Since D is the value of the trunk diameter in cm, a factor of 40 000 is used to obtain the area in m2, assuming the tree base has a circular form, in equation.


$$G=(π/40000)*D^2$$


```{r}
inventory$G=(pi/40000)*inventory$dbh^2
```



## Biomass
The aboveground biomass of individual trees (AGB) was estimated using two equations:Model 13, 33 (Marra et al, 2016)^[https://bg.copernicus.org/articles/13/1553/2016/] and Model 7(Chave et al, 2014)^[https://onlinelibrary.wiley.com/doi/10.1111/gcb.12629]

Equation 33 (Marra et al, 2016)
$$AGB= 0.230 D^{2.406}*(WD)^{0.88}$$
Equation 13 (Marra et al, 2016)
$$AGB= 0.183 D^{2.328}$$

Model 7 (Chave et al, 2014)
$$AGBest= exp [-1.803 - 0.97 E + 0.976 ln (WD) + 2.673 ln (D) - 0.229 [ln(D)]^2]$$
where:D=diameter (cm), WD=wood density (g/cm3)

### Write funtions for biomass

For Marra et al, 2016 (trees with wood density):
```{r}
biofun <- function(data, wd = data$wd, dbh = data$dbh) {
  #coeff
  b1 <- 0.230
  b2 <- 2.406
  b3 <- 0.880
  #all eq
  agb <- numeric(length(dbh))
  agb <- b1 * (dbh)^b2 * (wd)^b3
  return(agb)
}

```

For Marra et al, 2016 (trees without wood density):
```{r}
biofun2 <- function(data, dbh = data$dbh) {
  #coeff
  b1 <- 0.183
  b2 <- 2.328
  #all eq
  agb2 <- numeric(length(dbh))
  agb2 <- b1 * (dbh)^b2
  return(agb2)
}
 
```


For Chave et al, 2014:
```{r}
#variable E
ts <- 431 # temperature seasonality
cwd <- -39 # climate water deficit
ps <- 33 # precipitation seasonality
e <- (((0.178 * ts) - (0.938 * cwd) - (6.61 * ps)) * (10^-3))

#Model 7
ch7 <- function(data) {
  # Coefficients
  b1 <- -1.803
  b2 <- 0.976
  b3 <- 2.673
  b4 <- 0.0299
  # Equation
  agb_dry <- nrow(data)
  agb_dry <- b1 - (b2 * E) + (b2 * log(data$wd)) +
    (b3 * log(data$dbh)) - (b4 * (log(data$dbh1) ^ 2))
  agb_dry <- exp(agb_dry)
  h2o <- (agb_dry * 0.4827) # take of the H2O
  agb <- agb_dry + h2o
  return(agb)
}
```

### Aplpy funtions for biomass
```{r}
inventory$biomas=biofun2(inventory)

```


## 3D Visualization
3D Scatter plots are used to plot data points on three axes in the attempt to show the relationship between three forest variables. For more information check the 3D Rdocumentation ^[plot3d: 3D scatterplot, https://www.rdocumentation.org/packages/rgl/versions/0.107.14/topics/plot3d].


```{r, setup_rgl, message=FALSE, warning=FALSE}
library(rgl)
knitr::knit_hooks$set(webgl = hook_rgl)
```

```{r echo=TRUE, message=FALSE, warning=FALSE, rgl-plot, message=FALSE, webgl=TRUE}
plot3d(inventory$dbh,
       inventory$wd,
       xlab = "Diameter (cm)",
       ylab = "Heigth (m)",
       zlab = "Wood Density (g/cm3)",
       type = "s",
       size = 1,
       col = "red")
```
