# Chapter 2: Forest Composition & Structure

With our inventory data cleaned and standardized, we can now explore the ecological characteristics of our forest stand. Two of the most fundamental questions in any forest inventory are: What species are growing here? and How big are they?

In this chapter, we will use the ggplot2 package (which is loaded automatically with tidyverse) to create clear, publication-ready visualizations of our forest's composition and structure.

## Species Composition

Tropical and Neotropical forests are incredibly diverse. Visualizing the abundance of every single species in a single chart can result in a cluttered and unreadable mess. Instead, a standard practice is to isolate and visualize the most dominant species to understand the primary makeup of the canopy.

Let's calculate the top 10 most abundant species in our dataset and create a horizontal bar chart.
```{r}
#| label: species-composition
#| fig-cap: "Top 10 most abundant tree species in the inventory."

# Calculate species frequencies and filter for the top 10
top_species <- inventory_clean %>%
  count(Scientific_Name, name = "Count") %>%
  arrange(desc(Count)) %>%
  slice_max(order_by = Count, n = 10)

# Create a horizontal bar chart
ggplot(top_species, aes(x = reorder(Scientific_Name, Count), y = Count)) +
  geom_col(fill = "steelblue", color = "black", alpha = 0.8) +
  coord_flip() + # Flips the axes for easier reading of species names
  labs(
    title = "Dominant Species in the Forest Stand",
    x = "Scientific Name",
    y = "Number of Individuals"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(face = "italic") # Standard practice: italicize scientific names
  )
```

## Quantifying biodiversity: diversity indices
While our bar chart shows the most common species, it does not tell us the overall diversity of the stand. To quantify this, ecologists rely on diversity indices. The two most common are the Shannon-Wiener Index ($H'$) and the Simpson Index ($D$).

Shannon-Wiener Index ($H'$): Accounts for both abundance and evenness of the species present [@shannon1948; @magurran2004]. A higher value indicates higher diversity.The formula is:

$$H' = -\sum_{i=1}^{S} p_i \ln(p_i)$$
where $p_i$ is the proportion of individuals belonging to the $i$th species.

Simpson Index ($D$) [@simpson1949]: Measures the probability that two individuals randomly selected from a sample will belong to the same species.Let's use the vegan package to calculate these metrics for our plot.

```{r}
#| label: diversity-metrics
#| message: false

# Install vegan if needed: install.packages("vegan")
library(vegan)

# First, we need a table of species counts (a community matrix)
species_counts <- inventory_clean %>%
  mutate(Scientific_Name = ifelse(Scientific_Name == "", "Unknown", Scientific_Name)) %>%
  count(Scientific_Name) %>%
  pivot_wider(names_from = Scientific_Name, values_from = n, values_fill = 0)

# Calculate Shannon Index
shannon_index <- diversity(species_counts, index = "shannon")

# Calculate Simpson Index
simpson_index <- diversity(species_counts, index = "simpson")

# Calculate Species Richness (Total number of unique species)
richness <- specnumber(species_counts)

# Print the results
cat("Species Richness (S):", richness, "\n")
cat("Shannon Index (H'):", round(shannon_index, 3), "\n")
cat("Simpson Index (D):", round(simpson_index, 3), "\n")
```
## Species Accumulation Curves

In dense tropical canopies, a critical question is: Did we sample a large enough area to capture the true diversity of the forest?

A Species Accumulation Curve plots the number of new species discovered as the sampling effort (number of trees or plots) increases. If the curve flattens out (reaches an asymptote), our sampling effort was sufficient. If it is still climbing steeply, many rare species remain undiscovered.

```{r}
#| label: species-accumulation
#| fig-cap: "Species accumulation curve showing the rate at which new species are encountered as sampling effort increases."

presence_matrix <- inventory_clean %>%
  # 1. Add this filter to remove empty strings first!
  filter(Scientific_Name != "", !is.na(Scientific_Name)) %>%
  # 2. Proceed with the rest of the code
  mutate(Presence = 1, Tree_Index = row_number()) %>%
  pivot_wider(names_from = Scientific_Name, values_from = Presence, values_fill = 0) %>%
  select(-Plot_ID, -Tree_ID, -Common_Name, -DBH_cm, -Easting, -Northing, -Tree_Index)

accum_curve <- specaccum(presence_matrix, method = "exact")
plot(accum_curve, ci.type = "poly", col = "blue", lwd = 2, ci.lty = 0, ci.col = "lightblue",
     xlab = "Number of Trees Sampled",
     ylab = "Cumulative Number of Species",
     main = "Species Accumulation Curve")
```



```{r}
#| label: species-accumulation-subplot
#| fig-cap: "Species accumulation curve showing the cumulative discovery of species as the number of sampled subplots increases."
#| message: false
#| warning: false

library(vegan)
library(tibble) # Needed to convert the Plot_ID column into row names

# 1. Create a community matrix grouped by Subplot (Plot_ID)
subplot_matrix <- inventory_clean %>%
  # Ensure we have no blank species names
  filter(Scientific_Name != "", !is.na(Scientific_Name)) %>%
  # Count the number of each species PER subplot
  count(Plot_ID, Scientific_Name) %>%
  # Pivot into a wide matrix (Rows = Plot_ID, Columns = Species)
  pivot_wider(names_from = Scientific_Name, values_from = n, values_fill = 0) %>%
  # 'vegan' requires the matrix to be purely numeric, so we move Plot_ID to the row names
  column_to_rownames(var = "Plot_ID")

# 2. Calculate the accumulation curve using the subplots
accum_curve_subplot <- specaccum(subplot_matrix, method = "exact")

# 3. Plot the results
plot(accum_curve_subplot, 
     ci.type = "poly", col = "darkgreen", lwd = 2, ci.lty = 0, ci.col = "lightgreen",
     xlab = "Number of Subplots Sampled",
     ylab = "Cumulative Number of Species",
     main = "Species Accumulation Rate by Subplot")

# Add a grid for easier reading of the asymptote 
grid()
```

