# Chapter 2: Forest Composition & Structure

With our inventory data cleaned and standardized, we can now explore the ecological characteristics of our forest stand. Two of the most fundamental questions in any forest inventory are: What species are growing here? and How big are they?

In this chapter, we will use the ggplot2 package (which is loaded automatically with tidyverse) to create clear, publication-ready visualizations of our forest's composition and structure.

## 2.1 Species Composition

Tropical and Neotropical forests are incredibly diverse. Visualizing the abundance of every single species in a single chart can result in a cluttered and unreadable mess. Instead, a standard practice is to isolate and visualize the most dominant species to understand the primary makeup of the canopy.

Let's calculate the top 10 most abundant species in our dataset and create a horizontal bar chart.
```{r}
#| label: species-composition
#| fig-cap: "Top 10 most abundant tree species in the inventory."

# Calculate species frequencies and filter for the top 10
top_species <- inventory_clean %>%
  count(Scientific_Name, name = "Count") %>%
  arrange(desc(Count)) %>%
  slice_max(order_by = Count, n = 10)

# Create a horizontal bar chart
ggplot(top_species, aes(x = reorder(Scientific_Name, Count), y = Count)) +
  geom_col(fill = "steelblue", color = "black", alpha = 0.8) +
  coord_flip() + # Flips the axes for easier reading of species names
  labs(
    title = "Dominant Species in the Forest Stand",
    x = "Scientific Name",
    y = "Number of Individuals"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(face = "italic") # Standard practice: italicize scientific names
  )
```
## 2.2 Forest Stand Structure (DBH)

The distribution of tree diameters is a key indicator of forest health, history, and successional stage. In natural, uneven-aged forests, we typically expect to see a "Reverse-J" curve. This means there are many small, young trees in the understory and progressively fewer large, older trees forming the upper canopy.

We can visualize this using a histogram of our DBH_cm (Diameter at Breast Height) column.

```{r}
#| label: dbh-distribution
#| fig-cap: "Diameter distribution showing a classic Reverse-J curve for a natural forest."

ggplot(inventory_clean, aes(x = DBH_cm)) +
  # We use a binwidth of 5 cm, which is standard for forestry DBH classes
  geom_histogram(binwidth = 5, fill = "forestgreen", color = "black", alpha = 0.7) +
  # Create dynamic x-axis breaks based on our actual data max
  scale_x_continuous(breaks = seq(0, max(inventory_clean$DBH_cm, na.rm = TRUE), by = 10)) +
  labs(
    title = "Diameter Distribution of the Stand",
    subtitle = "The Reverse-J shape indicates an uneven-aged, regenerating forest",
    x = "Diameter at Breast Height (cm)",
    y = "Frequency (Number of Trees)"
  ) +
  theme_minimal()
```

