# Chapter 3: 2D Spatial Distribution and Mapping

Forestry is inherently spatial. Knowing what is in the forest and how big it is only tells half the story; we also need to know exactly where the trees are.

Thanks to the Easting (ESTE) and Northing (NORTE) coordinates collected in the field, we can map our forest stand. In this chapter, we will use the sf (Simple Features) package to convert our standard data table into a spatial object and ggplot2 to map it.

## Creating Spatial Data

Currently, our dataset is just a data frame (a table). To map it, R needs to understand that the Easting and Northing columns represent physical locations on the Earth.

We do this using the st_as_sf() function. We must also define the Coordinate Reference System (CRS). Since these are UTM coordinates (commonly used in South American forest inventories), we will assign an EPSG code. Note: You may need to adjust the EPSG code (e.g., 32718 for UTM Zone 18S) depending on the exact location of your study area.

```{r}
#| label: make-spatial
#| message: false

# Convert the cleaned data frame into an 'sf' spatial object
inventory_sf <- inventory_clean %>%
  st_as_sf(coords = c("Easting", "Northing"), 
           crs = 32718, # Replace with your specific UTM EPSG code if different
           remove = FALSE) # Keep original coordinate columns just in case

# Verify it is now a spatial object
class(inventory_sf)
```
## Mapping Forest Structure

Now that R knows these are spatial points, plotting them is as easy as adding geom_sf() to our standard ggplot2 workflow.

Let's start by plotting every tree in the stand. To make the map informative, we will scale the size of each point based on the tree's diameter (DBH_cm). This allows us to instantly locate the oldest, largest trees in the canopy.

```{r}
#| label: map-structure
#| fig-cap: "Spatial distribution of trees in the inventory plot. Point size correlates to tree diameter."

ggplot(data = inventory_sf) +
  # Use geom_sf for spatial data, mapping size to DBH
  geom_sf(aes(size = DBH_cm), color = "darkgreen", alpha = 0.6) +
  # Adjust the size scale so small trees are visible but large trees don't overlap too much
  scale_size_continuous(range = c(1, 6), name = "DBH (cm)") +
  labs(
    title = "Forest Stand Map: Tree Locations and Sizes",
    subtitle = "Larger circles represent trees with greater diameter at breast height",
    x = "Easting ",
    y = "Northing "
  ) +
  theme_minimal() +
  theme(
    # Angle the x-axis text so the long UTM coordinates don't overlap
    axis.text.x = element_text(angle = 45, hjust = 1) 
  )
```

## 3.3 Visualizing Species Distribution

Some tree species grow in clusters due to seed dispersal mechanisms or soil micro-conditions, while others are randomly scattered. We can visualize this by highlighting specific species on our map.

Looking at our dataset, Protium amazonicum (Copal) and species of the Eschweilera genus (like Machimango) are highly abundant. Let's create a map that specifically highlights the "Copal" trees against the rest of the stand.

```{r}
#| label: map-species
#| fig-cap: "Highlighting the spatial distribution of Copal (Protium amazonicum) within the stand."

ggplot(data = inventory_sf) +
  # Map color to whether the tree is Copal or not
  geom_sf(aes(size = DBH_cm, color = Common_Name == "Copal"), alpha = 0.7) +
  # Manually set colors: Grey for other trees, Red for Copal
  scale_color_manual(values = c("grey70", "firebrick"), 
                     labels = c("Other Species", "Copal"),
                     name = "Species") +
  scale_size_continuous(range = c(1, 5), guide = "none") + # Hide size legend to keep it clean
  labs(
    title = "Spatial Distribution of Copal (Protium amazonicum)",
    x = "Easting",
    y = "Northing"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## Analyzing Spatial Clustering with the Morisita Index

Visualizing points on a map is incredibly helpful, but our eyes can sometimes deceive us into seeing patterns that aren't statistically there. To objectively determine if a species is clustered, random, or uniformly distributed, we can calculate the Morisita Index of Dispersion (Id​).

The interpretation of the index is straightforward:

```{}
  $$Id​>1: Clustered (Aggregated) distribution.$$

    $$Id​=1: Random distribution.$$

    $$Id​<1: Uniform (Regular) distribution.$$

```
  
To calculate this, we need to divide our forest stand into equal-sized quadrats (a grid), count the number of individuals in each quadrat, and apply the Morisita formula. We will use the sf package to create our grid and base R to calculate the math for our most abundant species, Copal (Protium amazonicum).

```{r}
#| label: morisita-index
#| message: false
#| warning: false

# 1. Filter our spatial data for just the Copal trees
copal_sf <- inventory_sf %>%
  filter(Common_Name == "Copal")

# 2. Create a grid (quadrats) over our forest plot
# We will create a grid of roughly 20x20 meter squares based on the extent of our data
plot_grid <- st_make_grid(inventory_sf, n = c(5, 5)) # 5x5 grid (25 quadrats total)

# Convert the grid to an sf object so we can map and manipulate it
plot_grid_sf <- st_sf(quadrat_id = 1:length(plot_grid), geometry = plot_grid)

# 3. Count how many Copal trees fall into each quadrat
# st_intersects finds the trees in each grid cell; lengths() counts them
plot_grid_sf$tree_count <- lengths(st_intersects(plot_grid_sf, copal_sf))

# Let's quickly map our quadrats to see the counts!
ggplot() +
  geom_sf(data = plot_grid_sf, aes(fill = tree_count), alpha = 0.5, color = "black") +
  geom_sf(data = copal_sf, color = "red", size = 2) +
  scale_fill_viridis_c(name = "Copal Count") +
  labs(
    title = "Quadrat Counts for Copal",
    subtitle = "Red dots are individual Copal trees"
  ) +
  theme_minimal()

# 4. Calculate the Morisita Index mathematically
q <- nrow(plot_grid_sf)                     # Total number of quadrats (q)
N <- sum(plot_grid_sf$tree_count)           # Total number of Copal trees (N)
n <- plot_grid_sf$tree_count                # Number of trees in each quadrat (n_i)

# Morisita Formula: Id = q * [ sum(n * (n - 1)) / (N * (N - 1)) ]
morisita_index <- q * (sum(n * (n - 1)) / (N * (N - 1)))

# Print the result
cat("Total Quadrats (q):", q, "\n")
cat("Total Copal Trees (N):", N, "\n")
cat("Morisita Index (Id):", round(morisita_index, 3), "\n")

if(morisita_index > 1) {
  cat("Result: The distribution is CLUSTERED.")
} else if (morisita_index < 1) {
  cat("Result: The distribution is UNIFORM.")
} else {
  cat("Result: The distribution is RANDOM.")
}
```
## Spatial Visualization of DBH and Tree Height

While mapping species presence is crucial, foresters also need to understand the spatial distribution of biomass and canopy structure. Are the tallest trees clustered in a specific valley? Are the widest trees evenly distributed?

We can visualize multiple quantitative parameters simultaneously on a single map. Using our spatial object (inventory_sf),we will map the physical location of each tree, scale the size of the point by its DBH (DBH_cm), and apply a color gradient to represent its total height (Height_m).
```{r}
#| label: map-dbh-height
#| fig-cap: "Spatial distribution of forest structure. Point size represents DBH (cm) and color represents Total Height (m)."
#| warning: false

# Ensure Height is numeric (in case it imported as character)
inventory_sf <- inventory_sf %>%
  mutate(Height_m = as.numeric(Height_m)) %>%
  filter(!is.na(Height_m)) # Remove any rows without height data

# Create a multi-variable spatial plot
ggplot(data = inventory_sf) +
  # Map size to DBH, and color to Height
  geom_sf(aes(size = DBH_cm, color = Height_m), alpha = 0.8) +
  # Use a viridis color palette which is colorblind-friendly and intuitive for heights
  scale_color_viridis_c(option = "magma", name = "Total Height (m)") +
  # Scale the size of the points
  scale_size_continuous(range = c(1, 7), name = "DBH (cm)") +
  labs(
    title = "Forest Canopy Structure and Biomass Hotspots",
    subtitle = "Darker/Purple points are lower canopy; Brighter/Yellow points are emergent trees",
    x = "Easting",
    y = "Northing"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
```


## Forest Stand Structure (DBH)

The distribution of tree diameters is a key indicator of forest health, history, and successional stage. In natural, uneven-aged forests, we typically expect to see a "Reverse-J" curve. This means there are many small, young trees in the understory and progressively fewer large, older trees forming the upper canopy.

We can visualize this using a histogram of our DBH_cm (Diameter at Breast Height) column.

```{r}
#| label: dbh-distribution
#| fig-cap: "Diameter distribution showing a classic Reverse-J curve for a natural forest."

ggplot(inventory_clean, aes(x = DBH_cm)) +
  # We use a binwidth of 5 cm, which is standard for forestry DBH classes
  geom_histogram(binwidth = 5, fill = "forestgreen", color = "black", alpha = 0.7) +
  # Create dynamic x-axis breaks based on our actual data max
  scale_x_continuous(breaks = seq(0, max(inventory_clean$DBH_cm, na.rm = TRUE), by = 10)) +
  labs(
    title = "Diameter Distribution of the Stand",
    subtitle = "The Reverse-J shape indicates an uneven-aged, regenerating forest",
    x = "Diameter at Breast Height (cm)",
    y = "Frequency (Number of Trees)"
  ) +
  theme_minimal()
```


